---
title: "Download data"
author: 
- name: Sara Fonseca Costa
  affiliation:
  - &DEE Department of ecology and evolution, University of Lausanne, Switzerland
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
---

```{r include = FALSE}
knitr::opts_chunk$set(cache = FALSE, autodep = TRUE)
```

```{r load-libs, message = FALSE, warning = FALSE, echo=FALSE}
library(BgeeDB)
```

## All experiments

If you don't want to use filters, as showed in the previous section (Extract Information) where you target a particular experiment to download, you can always download all the data referent to the species using the `getData()` function.
Note that in the getData() you should specify which data type you would like to retrieve from Bgee by specifying that in the argument `dataType`. 

```{r download-data}
DrosMelRNASeq <- Bgee$new(species = "Drosophila_melanogaster", dataType = "rna_seq")
dataBgee_DM <- getData(DrosMelRNASeq)

## Number of experiments that exist in BgeeDB
length(dataBgee_DM)
```

## Subset of the data

In case you just have interest in download a particular experiment (as showed before), anatomical entity or developmental stage, you can just download this data by specifying that in the respective arguments: `experimentId` , `anatEntityId` and `stageId`. 

### Per experiment

Taking your experimenId result from the filtering in the previously section, download just this respective experiment by specifying that in the argument `experimentId` in the `getData()` function.

```{r download-subset}
DrosMelRNASeq_SRP002072 <- getData(DrosMelRNASeq, experimentId = "SRP002072")
head(DrosMelRNASeq_SRP002072)
```

### Specify an anatomical entity

You are also able to retrieve data from the database by specifying the anatomic entity of interest for this particular experiment. Here we will target adult organism as an example.

```{r download-subset-ann, eval=FALSE}
DrosMelRNASeq_annEnt <- getData(DrosMelRNASeq, experimentId = "SRP002072", anatEntityId = "UBERON:0007023")
head(DrosMelRNASeq_annEnt)
```

Note you can specify more than one anatomical entity in the `anatEntityId` argument.

```{r download-subset-ann-multi, eval=FALSE}
DrosMelRNASeq_annEnt_plus <- getData(DrosMelRNASeq, experimentId = "SRP002072", anatEntityId = c("UBERON:0007023","UBERON:0000922","UBERON:0002548"))
head(DrosMelRNASeq_annEnt_plus)
```

### Specify a developmental stage

As demonstrated before, for experiment and anatomical entities, you also can download data by specifying your target developmental stage of interest in the `stageId` argument.

```{r download-subset-stg, eval=FALSE}
DrosMelRNASeq_SRP002072_stg <- getData(DrosMelRNASeq, stageId = c("UBERON:0000068","FBdv:00005341"))
head(DrosMelRNASeq_SRP002072_stg)
```


