---
title: "Software & Data"
author: 
- name: Sara Fonseca Costa
  affiliation:
  - &DEE Department of ecology and evolution, University of Lausanne, Switzerland
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
---

## Software installation

For this course, you will need R.3.6, which you can download from https://cran.r-project.org/ (if you are using RStudio cloud, that is recommended for this course, this is not necessary) and Bioconductor software version 3.10.

First install BiocManager package, to help you to install and manage packages from Bioconductor project. The code below can be passed directly in the terminal.

```{r eval = FALSE, echo = TRUE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
    BiocManager::install(version = "3.10")
```

Then install the BgeeDB package that will be used for this course, you can downloaded the package from Bioconductor by typing:

```{r eval = FALSE, echo = TRUE}
BiocManager::install("BgeeDB")
```

Some additional packages are used during the course and in that case to install them you just have to type: 

```{r eval = FALSE, echo = TRUE}
BiocManager::install("PACKAGE_NAME")
```

Note: If you want, you can also install RStudio software, which can be download from https://rstudio.com/products/rstudio/download/ to run later your code in your local machine.


## Data

The data used during this course will be,

Model organism:

- *Drosophila melanogaster* (DM)

Non-model organism:

- *Bos Taurus* (BT)

In more details: the steps Extract information, Download data and Processing data we will use DM data.
For the TopAnat and Sparql endpoint we will use BT data.


## References

Komljenovic A, Roux J, Robinson-Rechavi M, and Bastian F (2016). BgeeDB, an R package for retrieval of curated expression datasets
  and for gene list enrichment tests. F1000Research

  Bastian F et al. (2008). Bgee: Integrating and Comparing Heterogeneous Transcriptome Data Among Species. Data Integration Life Sci.
  Lecture Notes in Computer Science, 5109, 124-31
  
  