---
title: "TopAnat"
author: 
- name: Sara Fonseca Costa
  affiliation:
  - &DEE Department of ecology and evolution, University of Lausanne & Swiss Institute of Bioinformatics, Switzerland
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the packages

```{r packages, message = FALSE, warning = FALSE}
library(BgeeDB)
library(biomaRt)
library(ggplot2)
library(here)
```

The TopAnat allows you to make GO-like enrichment of anatomical terms, mapped to genes by expression patterns.

In this section we will use 2 different species, *Drosophila melanogaster* and *Bos taurus* to run the TopAnat analysis.

Initially we will use the output results from our previous analysis with *Drosophila melanogaster* to run TopAnat, after that we will run the TopAnat tool with data retrieved from biomart that match specific phenotype from a non-model organism.

## Model organism

### Create a Bgee object

In order to start the analysis in TopAnat you have to create a Bgee object, as demonstrated before in the download data section.

Note that in this example we do not specify any data type in the argument dataType, which means we will use all data types from Bgee.

```{r bgee-object-DM}
targetDrosophila <- Bgee$new(species = "Drosophila_melanogaster")
```

### loadTopAnatData()

After that you should download the data for the reference species in order to allows you to perform the TopAnat analysis. The function loadTopAnatData() loads multiple files for the target species, as:

- an organ relationships file

- an organ names file

- a gene to organs mapping file

```{r loadTopAnat-DM}
drosophilaTopAnatData <- loadTopAnatData(targetDrosophila, confidence = "silver", stage = "UBERON:0000092")
```

Note that the loadTopAnatData() function allows you to retrieve data from a specific developmental stage by using the `stage` argument and allow you to specify the confidence of the quality of expression calls by using the argument `confidence`. In this particular case we use all data and the developmental stage UBERON:0000092.

### Background/foreground genes

The next step of the analysis relies on topGO package, for this a list of genes in the foreground and in the background are necessary.

Note that in this case the foreground genes will be the genes detected with high variability between the two samples from the experiment SRP002072 analysed in the processing data section.

### Retrieve info from biomaRt

The initial step is to retrieve information (ensembl gene id) for the target species by using biomart.

```{r biomart-DM}
ensemblDM <- useMart("ENSEMBL_MART_ENSEMBL", dataset="dmelanogaster_gene_ensembl", host="mar2016.archive.ensembl.org")

# background genes
universeDM <- getBM(attributes=c("ensembl_gene_id"),mart=ensemblDM)

# foreground genes from the analysis
foregroundDM <- read.table(file = here("genes_Drosophila_melanogaster.tsv"), header=F, sep="\t")
foregroundDM <- unlist(foregroundDM, use.names = FALSE)
```

We will use this list of genes to create our geneList object.

```{r geneList-DM}
geneListDM <- factor(as.integer(unique(universeDM$ensembl_gene_id) %in% foregroundDM))
names(geneListDM) <- unique(universeDM$ensembl_gene_id)
summary(geneListDM)
```

### topGO object

Create the object with data loaded and the gene list with foreground genes.

```{r topGO-object-DM}
myTopAnatObject_DM <-  topAnat(drosophilaTopAnatData, geneListDM)
```

### Enrichment test

After creating the final object, the enrichment analysis is performed by using the statistical tests implemented in topGO package.

```{r enrichment-test-DM}
results_DM <- runTest(myTopAnatObject_DM, algorithm = 'classic', statistic = 'fisher')
```

### Results

The results can be retrieved by specifying the desired cutoff using the `cutoff` argument. You can also export the table by ordering the desire column specifying that in `ordering` argument.

```{r results-DM}
tableOver <- makeTable(drosophilaTopAnatData, myTopAnatObject_DM, results_DM)
## look 
head(tableOver,4)
```


## Non-Model organism

In this example we will use a non-model organism *Bos taurus* (cattle) to run the TopAnat analysis. 

### Create a Bgee object

In order to start the analysis in TopAnat you have to create a new Bgee object, as demonstrated before.

Note that in this example we do not specify any data type in the argument dataType, as in the previous example, which means that we will use all data types from Bgee once again.

```{r bgee-object-BT}
targetBT <- Bgee$new(species = "Bos_taurus")
```

### loadTopAnatData()

After that you should download the data for the reference species in order to allows you to perform the TopAnat analysis. 
Note that in this case we will not specify any developmental stage or confidence in the calls.

```{r loadTopAnat-BT}
bostaurusTopAnatData <- loadTopAnatData(targetBT)
```

### Background/foreground genes

In the next step we will perform the TopAnat analysis with topGO, for this a list of genes in the foreground and in the background are necessary.

In this example we will retrieve genes enriched for expression in muscle. 

### Retrieve info from biomaRt

The initial step is to retrieve information for the target species by using biomart.

```{r biomart-BT}
ensembl_BT <- useMart("ENSEMBL_MART_ENSEMBL", dataset="btaurus_gene_ensembl", host="mar2016.archive.ensembl.org")

# background genes
universe_BT <- getBM(attributes=c("ensembl_gene_id","phenotype_description"),mart=ensembl_BT)

# select phenotypes related to muscle
phenotypes <- grep("musc", unique(universe_BT$phenotype_description), value=T)

# Foreground genes are those with an annotated phenotype related to muscle.
myGenes <- unique(universe_BT$ensembl_gene_id[universe_BT$phenotype_description %in% phenotypes])

# list vector with 0 and 1  
geneList_BT <- factor(as.integer(unique(universe_BT$ensembl_gene_id) %in% myGenes))
names(geneList_BT) <- unique(universe_BT$ensembl_gene_id)
summary(geneList_BT)
```

Note that you have just 7 genes in the foreground that match 'musc' keyword when you extract uniques with `getBM()` function.

If we do a search in UniProtKB/Swiss-Prot entries with keyword 'muscle' we will found the following genes that can be part of our foreground:

```{r genes-UniProt}
phenotypes <- c("ENSBTAG00000008394", "ENSBTAG00000006030", "ENSBTAG00000015644", "ENSBTAG00000011424", "ENSBTAG00000010799", "ENSBTAG00000014614", "ENSBTAG00000011869", "ENSBTAG00000045757", "ENSBTAG00000026266", "ENSBTAG00000022158", "ENSBTAG00000006419", "ENSBTAG00000030425", "ENSBTAG00000009707", "ENSBTAG00000009749", "ENSBTAG00000033217", "ENSBTAG00000007196", "ENSBTAG00000021218", "ENSBTAG00000005333", "ENSBTAG00000006424", "ENSBTAG00000005353", "ENSBTAG00000021992", "ENSBTAG00000026972", "ENSBTAG00000005970", "ENSBTAG00000046332", "ENSBTAG00000018285", "ENSBTAG00000005714", "ENSBTAG00000016024", "ENSBTAG00000017992", "ENSBTAG00000014003", "ENSBTAG00000018204", "ENSBTAG00000019159")
```

We will use this list of genes to create our geneList object.

```{r genes-UniProt-geneList}
geneList_BT <- factor(as.integer(unique(universe_BT$ensembl_gene_id) %in% phenotypes))
names(geneList_BT) <- unique(universe_BT$ensembl_gene_id)
summary(geneList_BT)
```

### topGO object

Create the object with data loaded and the gene list with foreground genes (phenotypes from the search in UniProtKB/Swiss-Prot).

```{r topGO-object-BT}
myTopAnatObject_BT <-  topAnat(bostaurusTopAnatData, geneList_BT)
```

### Enrichment test

After creating the final object, the enrichment analysis is performed by using the statistical tests implemented in topGO package.
Note that in this example we will use weight in the argument `algorithm`.

```{r enrichment-test}
results_BT <- runTest(myTopAnatObject_BT, algorithm = 'weight', statistic = 'fisher')
```

### Results

#### Global results

The results can be retrieved by specifying the desired cutoff using the `cutoff` argument. You can also export the table by ordering the desire column specifying that in `ordering` argument.


```{r global-results, fig.width=14,fig.height=5}
tableOver <- makeTable(bostaurusTopAnatData, myTopAnatObject_BT, results_BT)

grph <- ggplot(tableOver, aes(x = reorder(organId, FDR), y = FDR))
grph <- grph + geom_bar(stat="identity", color='black',fill='gray')
grph <- grph + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + xlab("Anatomic entities") + ylab("FDR") + ggtitle("Results - enrichment test")
grph
```

#### Retrieve significant genes

```{r small-FDR}
tableOver <- makeTable(bostaurusTopAnatData, myTopAnatObject_BT, results_BT, cutoff = 0.01, ordering = -8)

targetUberon <- c("UBERON:0001134", "UBERON:0001111", "UBERON:0001401", "UBERON:0002000", "UBERON:0034908")

termStat(myTopAnatObject_BT, targetUberon) 
```

In order to use some target gene in the next section (Sparql endpoint), we should retrieve the ensembl IDs for one of the anatomical entities that are significant.

```{r ensembl-ids}
annotated <- genesInTerm(myTopAnatObject_BT, targetUberon)[["UBERON:0001134"]]
annotated[annotated %in% sigGenes(myTopAnatObject_BT)]
```

